---
title: "Exploring Vaccine-Induced Immune Response"
author: "Luka Wamalwa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
``` 

## 🧪 Project Overview
This project simulates and analyzes a clinical dataset representing antibody responses following COVID-19 vaccination. It includes demographic, clinical, and immunological response features. Analysis includes ANOVA, logistic regression, and insightful visualizations using `ggpubr`, `ggpmisc`, and `gghighlight`.

---

## 1. Load Required Libraries
```{r}
required_packages <- c("tidyverse", "ggpubr", "ggsignif", "ggpmisc", "gghighlight", "broom", "readr")
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  library(pkg, character.only = TRUE)
}
```

---

## 2. Simulate Patient Demographics and Vaccine Data
```{r}
set.seed(123)
n <- 500
patients <- tibble(
  patient_id = 1:n,
  age = round(rnorm(n, 45, 15)),
  sex = sample(c("Male", "Female"), n, replace = TRUE),
  bmi = round(rnorm(n, 25, 4), 1),
  comorbidity = sample(c("None", "Hypertension", "Diabetes", "Both"),
                       n, replace = TRUE, prob = c(0.5, 0.2, 0.2, 0.1)),
  dose = sample(c("Dose 1", "Dose 2", "Booster"), n, replace = TRUE)
)
```

---

## 3. Simulate Longitudinal Antibody Titers
```{r}
timepoints <- c(0, 14, 28, 90)
sim_data <- expand.grid(patient_id = patients$patient_id, time = timepoints) %>%
  left_join(patients, by = "patient_id") %>%
  mutate(
    base_titer = rnorm(nrow(.), 20, 5),
    titer = base_titer +
      case_when(
        dose == "Dose 1" ~ time * 0.8,
        dose == "Dose 2" ~ time * 1.2,
        dose == "Booster" ~ time * 1.5
      ) +
      ifelse(comorbidity != "None", -runif(nrow(.), 2, 5), 0) +
      rnorm(nrow(.), 0, 10),
    seropositive = ifelse(titer > 50, 1, 0)
  )

if (!dir.exists("data")) dir.create("data")
write_csv(sim_data, "data/simulated_vaccine_data.csv")
```

---

## 4. Statistical Analysis (Day 28)
```{r}
day28 <- sim_data %>% filter(time == 28)

# ANOVA
anova_model <- aov(titer ~ dose, data = day28)
summary(anova_model)
TukeyHSD(anova_model)

# Logistic Regression
logit_model <- glm(seropositive ~ age + sex + bmi + comorbidity + dose,
                   data = day28, family = binomial)
summary(logit_model)
tidy(logit_model, exponentiate = TRUE, conf.int = TRUE)
```

---

## 5. Visualizations

### A. Boxplot with Significance using `ggpubr` and `ggsignif`
```{r}
ggboxplot(day28, x = "dose", y = "titer", color = "dose", palette = "jco", add = "jitter") +
  stat_compare_means(method = "anova", label.y = 120) +
  stat_compare_means(label = "p.signif", method = "t.test",
                     comparisons = list(c("Dose 1", "Dose 2"), c("Dose 2", "Booster"), c("Dose 1", "Booster")),
                     tip.length = 0.01) +
  labs(title = "Antibody Titers by Vaccine Dose at Day 28",
       y = "Antibody Titer", x = "Dose")
```

### B. Mean Titer Over Time + Poly Eq (using `ggpmisc`)
```{r}
mean_titers <- sim_data %>%
  group_by(dose, time) %>%
  summarise(mean_titer = mean(titer), .groups = "drop")

ggplot(mean_titers, aes(x = time, y = mean_titer, color = dose)) +
  geom_line(linewidth = 1.2) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2, raw = TRUE), se = FALSE) +
  stat_poly_eq(
    formula = y ~ poly(x, 2, raw = TRUE),
    aes(label = paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~")),
    parse = TRUE
  ) +
  labs(
    title = "Mean Antibody Response Over Time by Dose",
    x = "Days Since Vaccination", y = "Mean Antibody Titer"
  ) +
  theme_minimal()
```

### C. Top 10% Responders Highlight (using `gghighlight`)
```{r}
top90 <- sim_data %>% filter(time == 90)

ggplot(top90, aes(x = age, y = titer, color = dose)) +
  geom_point(size = 2, alpha = 0.6) +
  gghighlight(
    titer > quantile(titer, 0.9),
    label_key = dose,
    use_direct_label = TRUE
  ) +
  labs(
    title = "Top 10% Responders at Day 90",
    x = "Age", y = "Antibody Titer"
  ) +
  theme_minimal()
```

---

## 📌 Note on Model Fit
> The polynomial regression fits shown for each dose group are based on **mean antibody titers at four fixed timepoints**. Because of the deterministic structure of the simulation, these yield near-perfect adjusted R² values (≈ 1.00), which is expected and not misleading.

---

